---
title: "Introduction"
author: "Dustin Johnson"
date: "2/10/2017"
output: 
  html_document:
    keep_md: true
    theme: journal
---

For our first seminar, the following was covered (please refer to slides on connect):

1. Setting up and getting familiar with R and R Studio.
2. Reading and writing (IO)
3. Data structures in R (matrices, vectors, data.frames, lists)
4. Data types in R (numeric, integer, string/character, factor)
5. R logical operators (>, < , ==, etc.)
6. Examining data upon import
7. Extracting information and data from data.frames

We will walk through the process with an applied example using the Gapminder data set. Firstly, we are required to import the data. Conveniently, the data can be retrieved from [CRAN](https://cran.r-project.org/doc/FAQ/R-FAQ.html#What-is-CRAN_003f).

```{r}
# uncomment if you haven't installed gapminder yet.
#install.packages('gapminder')

library(gapminder)
```

We can examine the structure of the data using the built-in R functions below. I find `head()` and `str()` to be quite useful, as they capture nearly all the information regarding the data. Try to answer the following questions:

1. What are the dimensions of the data?
2. What is the data structure? (data.frame, matrix, etc.) 
3. What are the variable names and data types? (int, num, chr, etc)
4. Can you explain what the data is that you are attempting to investigate?

```{r,eval=FALSE}
head(gapminder); tail(gapminder)
nrow(gapminder); ncol(gapminder)
dim(gapminder)
```

```{r}
str(gapminder)
```

Once we know something about the data, we can begin to examine it more closely. Let's plot lifeExp from 1960 to 2007 for the country Rwanda. I'll leave it up to you to interpret.

```{r}
# rename object
gDat <- gapminder

# extract information - keep object names clear
# extract values from data.frame using [row index, column index]
# row index and column index could be names or integer values
# c() function create a vector that concatenates integers or names
# what does gDat$country == 'Rwanda' return?
gDat.lifeExp.Rwanda <- gDat[gDat$country == 'Rwanda',c('lifeExp', 'year')]
gDat.lifeExp.Rwanda.yearTrim <- gDat.lifeExp.Rwanda[gDat$year >= 1960,]

# plot the data
plot(lifeExp ~ year, 
		 data = gDat.lifeExp.Rwanda.yearTrim, 
		 type = "l", 
		 main = "Life Expectancy of Rwanda \n ~ 1962 to 2007 ~",
		 col = "blue")

```

Let's say we are interested in the rate of change of gdpPerCap for Rwanda We can apply functions to vectors and transform the data to the desired format and combine the results to the original data.frame using `cbind()` or `data.frame()`:

```{r}
# create a function to compute the rate of change
deltaCompute <- function(x)
{
	x.lag <- Hmisc::Lag(x) # what does the function lag() do?
	x.diff <- x - x.lag
	x.delta <- x.diff/x.lag

	return(x.delta)
}

# extract only the data we need
gDat.rwanda <- gDat[gDat$country == 'Rwanda',c('continent','year', 'gdpPercap')]

# use our function to compute the delta of gdpPercap
gDat.rwanda.delta <- deltaCompute(as.numeric(gDat.rwanda$gdpPercap))

# combine into original gDat.rwanda data.frame
gDat.rwanda.df <- data.frame(gDat.rwanda, Delta = gDat.rwanda.delta)

# plot results
plot(Delta ~ year, 
		 data = gDat.rwanda.df[-1,], 
		 main = "Annual Change of GDP per Capita for Rwanda",
		 type = "l",
		 col = "blue")

```
