---
title: "Data Wrangling"
author: "Dustin Johnson"
date: "2/10/2017"
output: 
  html_document:
    keep_md: true
    theme: journal
---


We begin to explore `data.frames` using simple functions that give us a brief understanding of what the data entails:

* `dim()`, `nrow()`, `ncol()`
* `str()`
* `is.data.frame()`, `is.matrix()`, `is.vector()`, etc.
* `summary()`

### dplyr <a name="dplyr"></a>

We then delved into the powerful dplyr package. Remember, `data.frames` are special forms of data objects in R that enable us to combine numerical, factor, string, and other types of data together. A simple matrix cannot combine different types of data, as we saw in class. We covered useful dplyr "verbs" that perform routine operations on your data, including:

* `select()` - select the variables (columns) of the data you are interested in.
* `filter()` - filter the rows of the data.frame utilising your logical operators `>`,`<`,`>=`,`<=`, and `==` .
* `group_by()` - group the data according to some category, for instance, group all data by continent.
* `mutate()` - mutate or transform one variable into another using some function. For example, create a new variable (column) that is a combination of two other variables (population and GDP).
* `summarise()`  - Summarise multiple values to a single value. For example, the mean GDP or total GDP of each continent, where the continent group was created using the function `group_by()`.

We also learned about the use of the piping commands `%>%`, using `command⌘+shift+m,` making data wrangling intuitive and easy. Here's a n example:

<br>

#### Challenge: Can you find the world growth of GDP per Capita in 1997? <a name="challenge_dplyr"></a>

I would just like to **select** the year, continent and gdpPercap variables from the gapminder data object, **mutate** gdpPercap into a new variable (column) called change (rate of change of gdpPercap), **filter** all years greater (but not equal to) 1952 (why?), **group_by year**, and **summarize** the data by the mean rate of change of gdpPercap per the grouped category, year.

```{r, echo=FALSE, message=FALSE}
library(plyr)
library(dplyr)
library(gapminder)
```

```{r}
gdp.delta <- gapminder %>%
    select(year, continent, gdpPercap) %>%
    mutate(change = 100*((gdpPercap - lag(gdpPercap)))/gdpPercap) %>%
    filter(year > 1952) %>%
    group_by(year) %>%
    summarise(gdp_delta = mean(change))
```

The result is an object `gdp.delta` with the mean rate of change of GDP per Capita for each year (5-year lag), as below:

```{r,echo=FALSE,results='asis'}
knitr::kable(gdp.delta, align = "c")
```